**Evaluation of machine learning methods for 16S rRNA gene data**

### 2019-03-04 submission/manuscript.Rmd

Advances in sequencing technology and decreasing costs of generating 16S rRNA gene sequences have allowed rapid exploration and taxonomic characterization of the human associated microbiome. Currently, the microbiome field is growing at an unprecedented rate and as a result, there is an increasing demand for reproducible methods to identify associations between members of the microbiome and human host. However, this is an undertaking as human associated microbial communities are remarkably uneven. It is unlikely that a single species can explain a phenotype. Instead, subsets of those communities, in relation to one another and to their host, account for the differences. Therefore, researchers have started to explore the utility of machine learning (ML) models that use microbiota associated biomarkers to predict a phenotype and to understand the microbial ecology of the host. However, currently the field's use of ML lacks clarity and consistency on which methods are used and how these methods are implemented. Moreover, there is a lack of consideration on why a particular ML model is utilized. Recently, there is a trend towards using more complex ML models [@galkin_human_2018; @reiman_using_2017; @fioravanti_phylogenetic_2017] without a discussion on if interpretibility is necessary for the study. If we want to generate reproducible and robust microbiota-associated biomarker models we need to (1) implement consistent, accessible and transparent machine learning practices; (2) use ML models that reflect the goal of the study as it will inform/define the expectations of model accuracy, complexity, interpretibility, computational efficiency and scalability.

One application of machine learning to microbiome data has been to classify patients as having colorectal tumors based on microbiota-associated biomarkers in the stool [@zeller_potential_2014; @zackular_human_2014; @baxter_dna_2016; @baxter_microbiota-based_2016; @hale_shifts_2017]. Colorectal cancer (CRC) is one of the leading causes of death in the US, therefore developing a non-invasive screening tool with ML is a remarkable innovation. However, afforementioned standards have not been established in these ML models. Here we performed an empirical analysis comparing several different modeling pipelines using the same dataset. We used a previously published CRC study [@baxter_dna_2016] which had fecal 16S rRNA gene sequences and immunochemical test (FIT) results from `r paste(nrow(shared))` patients. We built ML models using 16S abundances and FIT to predict patients with normal or screen relevant neoplasias (SRN) disease status. The study had `r paste(sum(sum(meta$dx=="normal")+sum(meta$Dx_Bin=="Adenoma")))` normal and `r paste(sum(sum(meta$dx=="cancer")+sum(meta$Dx_Bin=="adv Adenoma")))` SRN samples. We established modeling pipelines for L2-regularized logistic regression, L1 and L2 support vector machines (SVM) with linear and radial basis function kernels, a decision tree, random forest and XGBoost. These models increase in complexity while they decrease in interpretability. Our ML pipeline performed 100 data-splits and utilized held-out test data to evaluate generalization and predicton performance of each ML model. The mean AUROC varied from `r paste(round(test_all[performance_index[1],3], 2))` (std ± `r paste(round(test_all[performance_index[1],4], 2))`) to `r paste(round(test_all[performance_index[7],3], 2))` (std ± `r paste(round(test_all[performance_index[7],4], 2))`). Random Forest and XGBoost had the highest mean AUROC for detecting SRN. In terms of computational efficiency, L2 Logistic Regression trained the fastest (`r paste(round(walltime_df[walltime_index[1],2], 3))` hours, std ± `r paste(round(walltime_df[walltime_index[1],3], 3))`), while XGBoost took the longest (`r paste(round(walltime_df[walltime_index[6],2], 3))` hours, std ± `r paste(round(walltime_df[walltime_index[6],3], 3))`). We found that mean cross-validation and testing AUROC could vary by as much as `r paste(round(difference[difference_index[7]], 3))`, which highlights the importance of a seperate held-out test set for evaluation. Aside from evaluating generalization and classification performances for each of these models, this study established standards for modeling pipelines of microbiome-associated machine learning models.

@@

  The colonic disease status is re-defined as two encompassing classes; Normal or Screen Relevant Neoplasias (SRNs). Normal class includes patients with non-advanced adenomas or normal colons whereas SRN class includes patients with advanced adenomas or carcinomas. Colonic disease status is the label predicted with each classifier.   The bacterial abundances and FIT results are the features used to predict colonic disease status. Bacterial abundances are discrete data in the form of Operational Taxonomic Unit (OTU) counts. There are 6920 OTUs for each sample. FIT levels are continuous data present for each sample. Because the data are in different scales, Python programming language v, module scikit-learn v is used to transform features by scaling each feature to a [0-1] range (Table 1) [@scikit-learn].

>>>>

Advances in sequencing technology and decreasing costs of generating 16S rRNA gene sequences have allowed rapid exploration and taxonomic characterization of the human associated microbiome. Currently, the microbiome field is growing at an unprecedented rate and as a result, there is an increasing demand for reproducible methods to identify associations between members of the microbiome and human health. However, this is an undertaking as human associated microbial communities are remarkably uneven. It is unlikely that a single species can explain a disease. Instead, subsets of those communities, in relation to one another and to their host, account for the differences in the health outcomes. Therefore, researchers have started to explore the utility of machine learning (ML) models that use microbiota associated biomarkers to predict human health and to understand the microbial ecology of the human diseases such as liver cirrhosis, colorectal cancer, inflammatory bowel diseases (IBD), obesity, and type 2 diabetes [@zeller_potential_2014; @zackular_human_2014; @baxter_dna_2016; @baxter_microbiota-based_2016; @hale_shifts_2017; @pasolli_machine_2016; @sze_looking_2016; @walters_meta-analyses_2014; @vazquez-baeza_guiding_2018; @qin_alterations_2014; @geman_deep_2018]. However, currently the field's use of ML lacks clarity and consistency on which methods are used and how these methods are implemented. Moreover, there is a lack of deliberation on why a particular ML model is utilized. Recently, there is a trend towards using more complex ML models such as random forest, extreme gradient boosting and neural networks without a discussion on if and how much model interpretibility is necessary for the study [@galkin_human_2018; @reiman_using_2017; @fioravanti_phylogenetic_2017; @ @geman_deep_2018]. If we want to generate reproducible and robust microbiota-associated biomarker models we need to (1) implement consistent, accessible and transparent machine learning practices; (2) use ML models that reflect the goal of the study as it will inform the expectations of model accuracy, complexity, interpretibility, computational efficiency and scalability.

To shed light on how much task definition and modeling choices can affect the outcomes of ML studies, we performed an empirical analysis comparing several different modeling pipelines using the same dataset. We used a previously published colorectal cancer (CRC) study [@baxter_dna_2016] which had fecal 16S rRNA gene sequences and human hemoglobin concentrations from `r paste(nrow(shared))` patients. We built ML models using fecal 16S rRNA gene sequences and human hemoglobin concentrations to predict patients with normal or screen relevant neoplasias (SRN) disease status. The study had `r paste(sum(sum(meta$dx=="normal")+sum(meta$Dx_Bin=="Adenoma")))` normal and `r paste(sum(sum(meta$dx=="cancer")+sum(meta$Dx_Bin=="adv Adenoma")))` SRN samples. We established modeling pipelines for L2-regularized logistic regression, L1 and L2 support vector machines (SVM) with linear and radial basis function kernels, a decision tree, random forest and XGBoost. These models increase in complexity while they decrease in interpretability. Our ML pipeline performed 100 data-splits and utilized held-out test data to evaluate generalization and predicton performance of each ML model. The mean AUROC varied from `r paste(round(test_all[performance_index[1],3], 2))` (std ± `r paste(round(test_all[performance_index[1],4], 2))`) to `r paste(round(test_all[performance_index[7],3], 2))` (std ± `r paste(round(test_all[performance_index[7],4], 2))`). Random Forest and XGBoost had the highest mean AUROC for detecting SRN. Despite the simplicity, the L1-regularized linear kernel SVM performed on par with many of the non-linear models. In terms of computational efficiency, L2 Logistic Regression trained the fastest (`r paste(round(walltime_df[walltime_index[1],2], 3))` hours, std ± `r paste(round(walltime_df[walltime_index[1],3], 3))`), while XGBoost took the longest (`r paste(round(walltime_df[walltime_index[6],2], 3))` hours, std ± `r paste(round(walltime_df[walltime_index[6],3], 3))`). We found that mean cross-validation and testing AUROC could vary by as much as `r paste(round(difference[difference_index[7]], 3))`, which highlights the importance of a seperate held-out test set for evaluation. Aside from evaluating generalization and classification performances for each of these models, this study established standards for modeling pipelines of microbiome-associated machine learning models.

@@

  The colonic disease status is re-defined as two encompassing classes; Normal or Screen Relevant Neoplasias (SRNs). Normal class includes patients with non-advanced adenomas or normal colons whereas SRN class includes patients with advanced adenomas or carcinomas. Colonic disease status is the label predicted with each classifier.   The bacterial abundances and FIT results are the features used to predict colonic disease status. Bacterial abundances are discrete data in the form of Operational Taxonomic Unit (OTU) counts. There are 6920 OTUs for each sample. FIT levels are continuous data present for each sample. Because the data are in different scales, features are transformd by scaling each feature to a [0-1] range (Table 1).
